---
title: "Test1"
author: "Charlie"
date: "2/1/2022"
output:
  pdf_document: default
  html_document: default
  word_document: default
urlcolor: blue
---

```{=html}
<!-- This is a comment block in RMD (actually, it is HTML)
     You need to install MikTex on a PC or MacTex on a MAc to create PDF files.
     - MikTex installation: 
       set paper to Letter (not A4)
       set Install Missing PAckages on the Fly to Yes (not Ask Me First)
       keep automatical updates checkbox checked
     
     The first time you create a PDF file using KnitR will be very slow
       because MikTex is updating -- it will be much faster after that
       
-->
```
```{r echo=FALSE, fig.show='hide', message=FALSE, warning=FALSE, code=readLines("scripts/lesson05-ModifyMappedElements-new.R"), results="hide"}
# execute the script but suppress all output
# basically this puts the script in memory so it can used later
#source(file="lesson05-ModifyMappedElements-new.R");
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Newlines in RStudio are done by adding two spaces to the end of a line.

Things you can do in RMD:\
**Bold**\
*Italics*\
***Bold/Italics***

Headers must have a **blank line** before in order to work.

# Header 1

## Header 2

### Header 3

You can go up to Header 6... I think you see the pattern!

A hyperlink <https://unsplash.com/s/photos/llama>

More RMD features using the\
[RMD Cheat Sheat](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)

## Embedded R code {style="color:red; text-align: center;"}

[You can embed R code and output in this document using `r weatherData$avgTemp[1:10]` everything between the apostrophes is treated as script. So, you can also output the average temp `r mean(weatherData$avgTemp)`. You can use anything from your script!]{style="color: blue; text-align: center;"}

## Including Plots

You can also embed plots, for example:

```{r echo=FALSE}
plot(plot1)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
